---
title: "Predicting NBA Player Positions Using Decision Tree"
author: "Nicholas Burke"
date: "05 December 2019"
output: html_document
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

This project will use the computer program R to generate a decision tree to determine NBA player’s positions based upon various statistical categories. Data will be used from [basketballreference.com](https://www.basketball-reference.com/) to construct a classification tree. These types of trees can be used to sort categorical data appropriately, to classify player's position. As one of the following  

•	Point Guard

•	Shooting Guard

•	Small Forward

•	Power Forward

•	Center

Using these statistics that were tracked over the course of the season

•	Games played

•	 Games started

•	Minutes per game

•	 2pt & 3pt field goals/attempts

•	Free throws/attempts

•	Assists, turnovers

•	Rebounds

•	Fouls

•	Blocks

•	Steals

•	 Offensive and Defensive rating

To resolve the issue of discrepancies in playing time among players, we will use per 100 possessions statistics. The outcome of the decision tree will show the value of each category based on position the players. These values can be used to determine threshold certain players have to reach at their respective position.

Firstly in order to generate a decision tree which must have some data in order to input in the mechanism.  We will import data using the ballr package as well as other necessary dependencies. 


```{r, message=FALSE}
library (magrittr) 
library (ggplot2)
library (janitor) 
library (scales) 
library(ballr)
library(knitr)
library(ggplot2)
library(ggthemes)
library(dplyr)
library (rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
```


## NBA Statistics in R

We will now view the player statistics from the 2018-2019 NBA season. It is important to use 100 possessions statistics in order to make sure all players are measure equal to get a representative reading in the decision tree.

Below are some of the entries of the per 100 possessions per game statistics for players during the season.

```{r,message=FALSE}
per_100 <- NBAPerGameStatisticsPer100Poss(season = 2019)
kable(head(per_100), caption = "Per 100 Stats")
 
```

Let's examine the column names of this data frame using the following function

```{r}
colnames(per_100)
```


All 33 categories are not neccessarily useful in the construction of the decision tree. Such as the following variables

•	name

•	x

•	 link

•	rnk

•	team

They are of the categorical nature and too unique to each player. Also these variables

•	 games

•	games started

•	percentages

•	fouls



Will not be used be for we are only interested in players by positions and they do not add any extra value to our tree.

The following variables will be used in the decision tree 

•	fg, fga

•	x3p, x3pa

•	x2p, x2pa

•	ft, fta

•	trb

•	ast, tov

•	blk, stl

•	pts

We will construct a new data frame with the statistics of interest and to use going forward 
Below are some of the entries of our new data frame.

```{r}
P100 <- per_100 %>%
  select(pos,fg,fga,x3p,x3pa,x2p,x2pa,ft,fta,trb,ast,blk,stl,tov,pts)
kable(head (P100), caption = "Per 100 Stats")

```


It is vital for the decision tree to work, we must create a character string for each position.

•	"C"

•	"PF"

•	"SF"

•	"SG"

•	"PG" 

```{r}
per_100$pos= as.character(per_100$pos) 
```

Now we can begin to construct a visual for the decision tree, by using the package ‘rpart’.  With our new data frame we can construct a classification tree. 


```{r}
tree <- rpart(P100 ,data=per_100,method="class")
```

In order to verify which variables were used in making of the decision tree will call on the following function to see.

This will display the complexity parameter table for the fitted Rpart object.

```{r}
printcp(tree,digit=3)
```

This shows that the included variables are 

• ast

• blk

• stl

• trb

• x3pa 

will be used for the construction of the decision tree. The excluded variables are 

• fg, fga 

• x3p, x2p, x2pa

• ft, fta

• tov

• pts

## Decision Tree 

To plot the decision tree, we will use the rpart.plot library to plot the appropriate tree and label it accordingly.



```{r}
library(rpart.plot) 
plot(tree, uniform = TRUE, main="Position") 
text(tree, all=TRUE) 

```

Using the prp() function we can generate a better type of the same decision tree

```{r}
prp(tree) 
```


Using the fancyRpartplot() function we can generate a more fancy decision tree plot.


```{r}

fancyRpartPlot(tree, cex=0.5, sub = 'Decision Tree', type = 3) 

```

```{r}
rpart.plot(tree, box.palette="RdBu", shadow.col="gray", nn=TRUE)
```

Now for a more in depth look at the decision here are the calculation for each node and respective split in the decision tree.


```{r}
summary(tree)
```

# Explanation 


• The root node is the shooting guard position, the first split is total rebounds with the threshold being 9.8 rebounds. This is determined to be the best split of the data. This divides the data into two groups, if above threshold can be classified as centers and below threshold classified as shooting guards. 

• The next nodes classifies centers from shooting guards by 3 point attempts at 0.95 which leads to a terminal node for the below value for centers. 

• The decision node of 5.9 assists classifies point guards and shooting guards which leads to another terminal node for the above value for point guards.

• To classify shooting guards from power forward the determining factor is 1.5 blocks leading to a terminal node for the above value for power forwards. 

• The decision node of 7 rebounds separates small forwards and shooting guard leading to a terminal node for the below value for shooting guards. 

• The last decision nodes are for total rebounds of 12 for above that point for centers and below for power forwards. The other decision node for steals of 1.1 with below for power forwards and above for small forwards.

## Conclusion

Each statistical category can be used to determine the position in which the player will fall into based on total rebounds, 3 point attempts, assists, blocks and steals.  Each decision nodes represent the signification of that category and bench mark for that position. This classification tree can be used to identify players and help team determine what player’s goal and focus should be in certain categories. The classification tree can be extended to a team based model of respective statistics or even narrowed down to a specific team or sets of players.


